---
title: "MEGA study — predictive and multivariate modelling"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: kable
    theme: readable
    highlight: tango
---

```{r}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  comment = "#>"
)
```

Code and results for the classification and multi-omics models in the MEGA
study. Each section gives one analysis and the result it produces. Model
fitting is shown but not re-executed on render; results are read back from the
saved object named in the chunk beneath each model.

All classification analyses use responder status as the outcome and are
restricted to the 57 trained participants.

```{r}
library(tidyverse)
library(pROC)
library(caret)
library(knitr)
library(mixOmics)
```


# DIABLO multi-omics integration, POST

Three blocks: dietary intake, microbiota genus relative abundance, and serum
markers (zonulin, irisin, nine cytokines, faecal SCFAs, blood and plasma volume
measures, blood lactate and endotoxin). The serum block is named `cytokines` in
the code.

keepX is tuned and each block's AUC is computed with `auroc` on the same 57
participants used to fit the model, so the AUCs describe model fit rather than
out-of-sample performance and are exploratory. This applies to all three DIABLO
models.

```{r, eval = FALSE}
#### Blocks ####

X <- list(
  diet = as.matrix(metadata[dietary_names]),
  microbiome = as.matrix(taxa[taxa_names]),
  cytokines = as.matrix(metadata[biomarker_cols])
)
Y <- factor(metadata$Responder, levels = c("No", "Yes"))

design <- matrix(0.3, ncol = 3, nrow = 3, dimnames = list(names(X), names(X)))
diag(design) <- 0

NCOMP <- 2
KEEPX_GRID <- 3:15


#### Feature tuning ####

set.seed(123)
tune_full <- tune.block.splsda(
  X = X,
  Y = Y,
  ncomp = NCOMP,
  test.keepX = list(diet = KEEPX_GRID, microbiome = KEEPX_GRID, cytokines = KEEPX_GRID),
  design = design,
  validation = "Mfold",
  folds = 3,
  nrepeat = 5,
  dist = "centroids.dist",
  BPPARAM = BPP
)
keepX <- tune_full$choice.keepX


#### Model ####

final.diablo.model <- block.splsda(
  X = X,
  Y = Y,
  ncomp = NCOMP,
  keepX = keepX,
  design = design
)

auroc(final.diablo.model, roc.block = "microbiome", roc.comp = NCOMP, print = FALSE)
```

```{r}
a_post <- readRDS("results/diablo_post.rds")
auc_post <- auroc(a_post$model, roc.comp = 2, print = FALSE, plot = FALSE)

tibble(Block = names(auc_post), `Fitted AUC` = map_dbl(auc_post, ~ .x$comp2[1, "AUC"])) %>%
  kable(digits = 3)
```

```{r}
a_post$keepX
```


# DIABLO multi-omics integration, PRE

The same blocks, tuning and model, with the timepoint filter set to PRE.

```{r, eval = FALSE}
metadata <- metadata %>% dplyr::filter(PRE_POST == "PRE")

OTU_table <- physeq %>%
  tax_fix() %>%
  ps_filter(Group != "CT" & PRE_POST == "PRE") %>%
  tax_filter(min_prevalence = 1/10, undetected = 100, tax_level = "Genus") %>%
  tax_agg("Genus") %>%
  tax_transform("compositional", zero_replace = "halfmin", chain = TRUE) %>%
  tax_transform("log", zero_replace = "halfmin", chain = TRUE) %>%
  ps_melt()
```

```{r}
a_pre <- readRDS("results/diablo_pre.rds")
auc_pre <- auroc(a_pre$model, roc.comp = 2, print = FALSE, plot = FALSE)

tibble(Block = names(auc_pre), `Fitted AUC` = map_dbl(auc_pre, ~ .x$comp2[1, "AUC"])) %>%
  kable(digits = 3)
```

```{r}
a_pre$keepX
```


# DIABLO multi-omics integration, PRE-to-POST fold change

Both timepoints are carried through the same transformation, then collapsed to
one row per subject. The fold change is the POST value divided by the PRE
value of the log-transformed abundance.

```{r, eval = FALSE}
#### Blocks, both timepoints retained ####

OTU_table <- physeq %>%
  tax_fix() %>%
  ps_filter(Group != "CT") %>%
  tax_filter(min_prevalence = 1/10, undetected = 100, tax_level = "Genus") %>%
  tax_agg("Genus") %>%
  tax_transform("compositional", zero_replace = "halfmin", chain = TRUE) %>%
  tax_transform("log", zero_replace = "halfmin", chain = TRUE) %>%
  ps_melt()

taxa_fc <- taxa %>%
  dplyr::arrange(MEGA_ID, PRE_POST) %>%
  dplyr::group_by(MEGA_ID) %>%
  dplyr::summarise(
    dplyr::across(all_of(taxa_names), ~ .x[2] / .x[1]),
    .groups = "drop"
  )
```

```{r}
d_fc <- readRDS("results/diablo_foldchange.rds")
auc_fc <- auroc(d_fc$model, roc.comp = 2, print = FALSE, plot = FALSE)

tibble(Block = names(auc_fc), `Fitted AUC` = map_dbl(auc_fc, ~ .x$comp2[1, "AUC"])) %>%
  kable(digits = 3)
```

```{r}
d_fc$keepX
```


# POST logistic regression

Penalised logistic regression (elastic net via glmnet) on log10(x + 1)
transformed genus abundances at POST, 10% prevalence filter, tuned by 10-fold
cross-validation repeated 10 times. Because the genus ranking comes from a
model fitted to all participants, performance is estimated by repeating the
selection inside each of 100 outer cross-validation folds.

```{r, eval = FALSE}
ps_in <- physeq %>%
  tax_fix() %>%
  ps_filter(Group != "CT" & PRE_POST == "POST") %>%
  tax_filter(undetected = 100, min_prevalence = 1/10, tax_level = "Genus")

set.seed(123)
mm_lr <- microbiomeMarker::run_sl(
  ps_in,
  group = "Responder",
  nfolds = 10,
  nrepeats = 10,
  taxa_rank = "Genus",
  norm = "none",
  top_n = 15,
  transform = "log10p",
  method = "LR"
)
```

```{r}
lr <- readRDS("results/logistic_regression.rds")

tibble(Metric = "Nested cross-validated AUC", Value = lr$auc[["nested"]]) %>%
  kable(digits = 3)
```

The fifteen highest-importance genera.

```{r}
mk <- as.data.frame(microbiomeMarker::marker_table(lr$mm_lr))

tibble(
  Genus = as.character(mk$feature),
  `Enriched in` = as.character(mk$enrich_group),
  Importance = as.numeric(mk$ef_imp)
) %>%
  kable(digits = 2)
```


# PICRUSt2 pathway random forest, POST

Random forest on MetaCyc pathway abundances at POST, with age and sex as
additional predictors. Zeros are replaced with half of each pathway's minimum
non-zero value and abundances are log(1 + x) transformed. Zero-variance and
near-zero-variance pathways are removed within each resampling fold, and `mtry`
is tuned over five values. AUCs are computed with the direction of comparison
fixed to the responder class.

```{r, eval = FALSE}
rec <- recipe(Responder ~ ., data = rf_data_full) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors())

set.seed(123)
folds <- createMultiFolds(rf_data_full$Responder, k = 5, times = 3)
ctrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  index = folds,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

set.seed(123)
rf_clean <- train(
  rec,
  data = rf_data_full,
  method = "rf",
  trControl = ctrl,
  metric = "ROC",
  importance = TRUE,
  tuneLength = 5
)
```

```{r}
rf_post <- readRDS("results/random_forest_post.rds")
pred_post <- rf_post$clean$pred %>% dplyr::filter(mtry == rf_post$clean$bestTune$mtry)

tibble(
  Metric = c("Cross-validated AUC", "mtry"),
  Value = c(
    as.numeric(auc(roc(pred_post$obs, pred_post$Yes, levels = c("No", "Yes"), direction = "<", quiet = TRUE))),
    rf_post$clean$bestTune$mtry
  )
) %>%
  kable(digits = 3)
```

```{r}
read_csv("results/random_forest_post_importance.csv", show_col_types = FALSE) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::select(Pathway = variable, Importance = importance) %>%
  kable(digits = 2)
```


# PICRUSt2 pathway random forest, PRE-to-POST

Random forest on MetaCyc pathway abundances, PRE-to-POST ratio, with age and
sex as additional predictors. Zeros are replaced with half of each pathway's
minimum non-zero value and abundances are log(1 + x) transformed. Zero-variance,
near-zero-variance and highly correlated pathways are removed within each
resampling fold, and `mtry` is tuned over five values.

```{r, eval = FALSE}
rec <- recipe(Responder ~ ., data = rf_data_full) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_corr(all_numeric_predictors(), -Age, threshold = 0.9)

set.seed(123)
folds <- createMultiFolds(rf_data_full$Responder, k = 5, times = 3)
ctrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  index = folds,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

set.seed(123)
rf_clean <- train(
  rec,
  data = rf_data_full,
  method = "rf",
  trControl = ctrl,
  metric = "ROC",
  importance = TRUE,
  tuneLength = 5
)
```

```{r}
rf <- readRDS("results/random_forest.rds")
pred_clean <- rf$clean$pred %>% dplyr::filter(mtry == rf$clean$bestTune$mtry)

tibble(
  Metric = c("Cross-validated AUC", "mtry"),
  Value = c(
    as.numeric(auc(roc(pred_clean$obs, pred_clean$Yes, levels = c("No", "Yes"), direction = "<", quiet = TRUE))),
    rf$clean$bestTune$mtry
  )
) %>%
  kable(digits = 3)
```

```{r}
read_csv("results/random_forest_importance.csv", show_col_types = FALSE) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::select(Pathway = variable, Importance = importance) %>%
  kable(digits = 2)
```


# selbal balance

selbal on POST genus abundances, 10% prevalence filter, maximum of 12 taxa,
deviance as the selection criterion. The input is already log-transformed, so
selbal's own log transformation is switched off (`logt = FALSE`).

```{r, eval = FALSE}
combined <- physeq %>%
  tax_fix() %>%
  tax_filter(min_prevalence = 1/10, undetected = 100, tax_level = "Genus") %>%
  ps_filter(PRE_POST == "POST" & Group != "CT") %>%
  tax_agg("Genus") %>%
  tax_transform("compositional", zero_replace = "halfmin", chain = TRUE) %>%
  tax_transform("log", zero_replace = 1, chain = TRUE) %>%
  ps_melt()

set.seed(123)
sb <- selbal(
  x = as.matrix(combined[microbe_names]),
  y = if_else(combined$Responder == "Yes", 1, 0),
  maxV = 12,
  logit.acc = "Dev",
  draw = FALSE,
  logt = FALSE
)
```

```{r}
sb <- readRDS("results/selbal_balance.rds")

tibble(
  Side = c(rep("Numerator", length(sb$numerator)), rep("Denominator", length(sb$denominator))),
  Taxon = c(sb$numerator, sb$denominator)
) %>%
  kable()
```

# Fast-and-frugal trees

FFTrees fitted separately at PRE and POST to log-transformed genus abundances
with age, BMI, mHEI score and sex, 15% prevalence filter. Participants are
split once into training and test halves; cues and thresholds are selected on
the training half.

```{r, eval = FALSE}
mat <- physeq %>%
  tax_fix() %>%
  ps_filter(Group != "CT" & PRE_POST == "PRE") %>%
  tax_filter(undetected = 100, min_prevalence = 1.5/10, tax_level = "Genus")

set.seed(123)
train_ind <- caret::createDataPartition(mat$Responder, p = 0.5, list = FALSE)

pub <- FFTrees(
  Responder ~ .,
  data = mat[train_ind, ],
  data.test = mat[-train_ind, ],
  quiet = TRUE
)
```

PRE tree.

```{r}
readRDS("results/fft_pre.rds")$published %>%
  dplyr::select(Level = level, Cue = cue, Direction = dir, Threshold = thr) %>%
  kable()
```

POST tree.

```{r}
readRDS("results/fft_post.rds")$published %>%
  dplyr::select(Level = level, Cue = cue, Direction = dir, Threshold = thr) %>%
  kable()
```


# Session information

```{r}
sessionInfo()
```
